---
output: 
  github_document
---

```{r}
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/utils/check_covariances.R')
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/models/eki_normal.R')
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/samples/samples_normal.R')
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/utils/eki_helper.R')
```

```{r}
plot_covariance_against_sigma <- function(particle_func, data_func, 
                                          covariance_func, covariance_type, column,
                                          correlation = F) {
  sigma_seq <- seq(-2, 1, length.out = 10)
  res <- c()
  for (sigma in sigma_seq) {
    prior_params <- list(din.sd = 2,
                         phi.mean = 0, phi.sd = 1,
                         eta0.mean = 0, eta0.sd = 1,
                         sigma.mean = sigma, sigma.sd = 1)
    particles <- particle_func(50, prior_params)
    true_params <- list(sigma = 0.5, phi = 0.25, eta0 = 0.11, d_in = 0.5)
    true_unconstrained_params <- unconstrain_malaria_params(true_params)
    likelihood_samples <- data_func(50, particles, true_unconstrained_params)
    # print(mean(likelihood_samples))
    covariances <- calculate_covariances(particles, likelihood_samples, 
                                         correlation = correlation)
    res <- c(res, covariance_func(covariances, column))
  }
  print(res)
  plot(exp(sigma_seq), res, xlab = expression(sigma), ylab = 'Covariance', 
       main = glue('Empirical {covariance_type}'), cex.main = 2, cex.lab = 1.5)
  abline(h = 0, col = 'red')
}

plot_data_diff_against_sigma <- function(true_data, particle_func, 
                                        data_func, column) {
  
  d_y <- length(true_data)
  # I'm replicating this data for the number of particles to make the dimensions     easier to work with
  simulated_data <- matrix(true_data, nrow = 50, ncol = d_y, byrow = T)
  
  sigma_seq <- seq(-2, 1, length.out = 10)
  res <- c()
  
  for (sigma in sigma_seq) {
    prior_params <- list(din.sd = 2,
                         phi.mean = 0, phi.sd = 1,
                         eta0.mean = 0, eta0.sd = 1,
                         sigma.mean = sigma, sigma.sd = 1)
    particles <- particle_func(50, prior_params)
    true_params <- list(sigma = 0.5, phi = 0.25, eta0 = 0.11, d_in = 0.5)
    true_unconstrained_params <- unconstrain_malaria_params(true_params)
    likelihood_samples <- data_func(50, particles, true_unconstrained_params)
    # print(mean(likelihood_samples))
    res <- c(res, mean(simulated_data - likelihood_samples))
    
  }
  print(res)
  plot(exp(sigma_seq + 1/2), res, xlab = expression(sigma), ylab = 'Covariance', 
       main = 'Movement of Particles', cex.main = 2, cex.lab = 1.5)
  abline(h = 0, col = 'red')
    
  
}

plot_movement_against_sigma <- function(true_data, particle_func, 
                                        data_func, column) {
  
  d_y <- length(true_data)
  # I'm replicating this data for the number of particles to make the dimensions     easier to work with
  simulated_data <- matrix(true_data, nrow = 50, ncol = d_y, byrow = T)
  
  sigma_seq <- seq(-2, 1, length.out = 10)
  res <- c()
  
  for (sigma in sigma_seq) {
    prior_params <- list(din.sd = 2,
                         phi.mean = 0, phi.sd = 1,
                         eta0.mean = 0, eta0.sd = 1,
                         sigma.mean = sigma, sigma.sd = 1)
    particles <- particle_func(50, prior_params)
    true_params <- list(sigma = 0.5, phi = 0.25, eta0 = 0.11, d_in = 0.5)
    true_unconstrained_params <- unconstrain_malaria_params(true_params)
    likelihood_samples <- data_func(50, particles, true_unconstrained_params)
    # print(mean(likelihood_samples))
    covariances <- calculate_covariances(particles, likelihood_samples)
    new_particles <- update_particles(1, particles, simulated_data, 
                                  likelihood_samples, covariances, 50)
    update_size <- new_particles - particles
    res <- c(res, mean(update_size[, column]))
    
  }
  print(res)
  plot(exp(sigma_seq), res, xlab = expression(sigma), ylab = 'Covariance', 
       main = 'Movement of Particles', cex.main = 2, cex.lab = 1.5)
  abline(h = 0, col = 'red')
    
  
}
```

# Checking Covariances

When testing the EKI algorithm on a multivariate normal model with unknown mean and noise, it was found that the EKI algorithm was not able to identify the true noise parameter, and was anchored to the prior distribution. The purpose of this markdown document is to explain this phenomenon.

The main functions for this document can be found in `src/utils/check_covariances.R` I am interested in looking at three variables:

-   How a covariance matrix changes with the prior mean of $\sigma^2$: `plot_covariance_against_sigma2`

-   How a covariance matrix changes with the prior variance of $\sigma^2$: `plot_covariance_against_sigma2_dispersion`

-   How a covariance matrix changes with the prior mean of $\alpha$: `plot_covariance_against_alpha`

# Findings

## Checking $C^{yy}$

In this section, I check the covariance of the generated data. In the known variance case the large sample limit is $HQH^T + R = 25x^Tx + \sigma^2I$. Thus, we anticipate that the diagonals of the covariance matrix will increase as the prior mean of the $p(\sigma^2)$ increases. Additionally, we expect no dependence on the mean parameter $\alpha$.

```{r}
plot_covariance_against_sigma(initialise_malaria_func, synthetic_malaria_func, 
                              check_c_yy, "Data Variance")
```

The results are what we expect, where there is a strong positive linear relationship between the mean of the diagonals and $\sigma^2$ but no relationship between the mean of the diagonals and $\alpha$.

## Checking $C^{yx}$

In this section, I check the cross covariance of the generated data and the current particles. In the known variance case the large sample limit is $HQ = 25x^T$. Of course, this is a $d_y \times 1$ matrix, but the cross covariance in the unknown variance case will be $d_y \times 2$. I conjecture that the second column will be zeros, since there is no correlation between the noise parameter $\sigma^2$ and the generated data. The intuition is that higher levels of $\sigma^2$ increase the dispersion of the generated data but not its location. To test for this I look at the mean of the entries in the second column of the covariance matrices.

```{r}
plot_covariance_against_sigma(initialise_malaria_func, synthetic_malaria_func, 
                              check_c_yx, "cross covariance", 1)
plot_covariance_against_sigma(initialise_malaria_func, synthetic_malaria_func, 
                              check_c_yx, "cross covariance", 2)
plot_covariance_against_sigma(initialise_malaria_func, synthetic_malaria_func, 
                              check_c_yx, "cross covariance", 3)
plot_covariance_against_sigma(initialise_malaria_func, synthetic_malaria_func,
                              check_c_yx, "cross covariance", 4)
```

In all plots we see a random fluctuation around zero. Interestingly, as the dispersion of the prior for $\sigma^2$ increases, so does the magnitude of these fluctuations. Ideally, we would want to see this same behavior reflected in the step-size, since the less confident we are in the prior, the more our particles should respond to the data.

```{r}
plot_covariance_against_sigma(initialise_malaria_func, synthetic_malaria_func, 
                              check_c_yx, "cross correlation", 1, 
                              correlation = T)
plot_covariance_against_sigma(initialise_malaria_func, synthetic_malaria_func, 
                              check_c_yx, "cross correlation", 2, 
                              correlation = T)
plot_covariance_against_sigma(initialise_malaria_func, synthetic_malaria_func, 
                              check_c_yx, "cross correlation", 3,
                              correlation = T)
plot_covariance_against_sigma(initialise_malaria_func, synthetic_malaria_func,
                              check_c_yx, "cross correlation", 4,
                              correlation = T)
```

## Checking Stepsize

I am also interested in looking at how the particle movement for $\sigma^2$ changes as I vary the parameters of the model. Specifically I am interested in the second column of the matrix

$C^{xy}(C^{yy})^{-1}$ which is equivalent to $C^{xy}(C^{yy} +(h_l^{-1}-1)C^{y|x})^{-1}$ when $h_l = 1$ (i.e. no tempering)

If the cross covariance matrix contains zeros, then we expect the step-size to theoretically be zero as well (empirically very small), and thus the particles won't move much from their prior distribution.

```{r}
plot_covariance_against_sigma(initialise_malaria_func, synthetic_malaria_func, 
                              check_stepsize, "Stepsize", 1)
plot_covariance_against_sigma(initialise_malaria_func, synthetic_malaria_func,
                              check_stepsize, "Stepsize", 2)
plot_covariance_against_sigma(initialise_malaria_func, synthetic_malaria_func,
                              check_stepsize, "Stepsize", 3)
plot_covariance_against_sigma(initialise_malaria_func, synthetic_malaria_func,
                              check_stepsize, "Stepsize", 4)
```

We can see that the step-size appears to fluctuate randomly around zero, and the magnitude of these fluctuations is extremely small. It is worth noting that this is without tempering - if we employ tempering the magnitude of each step will be even smaller.

Unlike $C^{yx}$ the magnitude of the step-size does not appear to change as the prior variance increases. I believe this is due to offsetting effects between the components $(C^{yy} + C^{y|x})^{-1}$ and $C^{yx}$. All else equal, as the prior variances increases $E[\sigma^2]$ increases. This leads to an increase in $C^{yx}$ but a decrease in $(C^{yy} + C^{y|x})^{-1}$.

## Checking Data Difference

```{r}
true_params <- list(sigma = 0.5, phi = 0.25, eta0 = 0.11, d_in = 0.5)
true_unconstrained_params = unconstrain_malaria_params(true_params)
true_data <- likelihood_malaria(true_unconstrained_params, log_obs = F)
plot_data_diff_against_sigma(true_data, initialise_malaria_func, 
                            synthetic_malaria_func, 1)
plot_data_diff_against_sigma(true_data, initialise_malaria_func, 
                            synthetic_malaria_func, 2)
plot_data_diff_against_sigma(true_data, initialise_malaria_func, 
                            synthetic_malaria_func, 3)
plot_data_diff_against_sigma(true_data, initialise_malaria_func, 
                            synthetic_malaria_func, 4)
```

## Checking Update Size

```{r}
true_params <- list(sigma = 0.5, phi = 0.25, eta0 = 0.11, d_in = 0.5)
true_unconstrained_params = unconstrain_malaria_params(true_params)
true_data <- likelihood_malaria(true_unconstrained_params, log_obs = F)
plot_movement_against_sigma(true_data, initialise_malaria_func, 
                            synthetic_malaria_func, 1)
plot_movement_against_sigma(true_data, initialise_malaria_func, 
                            synthetic_malaria_func, 2)
plot_movement_against_sigma(true_data, initialise_malaria_func, 
                            synthetic_malaria_func, 3)
plot_movement_against_sigma(true_data, initialise_malaria_func, 
                            synthetic_malaria_func, 4)
```

# Implications
