```{r}
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/eki_normal_known_var.R')
```

```{r}
plot_eki_normal_known_var <- function(eki_result, parameters) {
  
  alpha_particles <- eki_result$particles[, 1]
  true_alpha <- parameters$alpha
  
  par(mfrow = c(1, 1))
  
  hist(alpha_particles, freq = F, main = 'Alpha', xlab = '')
  abline(v = true_alpha, col = 'red')
}
```

# Results For Multivariate Normal Model

## The Model

We assume that we have access to a single observation from a multivariate normal distribution $y \sim N(\alpha x, 2^2I)$ where $\alpha$ is a scalar and $x$ is a known vector. The unknown parameter in this model is $\alpha$.

I look at a range of combinations of $\alpha, x$ and $\sigma^2$, comparing against MCMC sampling.

In all combinations we draw from the same prior:

-   $\alpha \sim ~ N(0, 5^2)$

## Changing Number of Dimensions

In this experiment I keep $\alpha$ and $\sigma^2$ fixed at 2 and 4 respectively and let $x$ be a vector of 1s corresponding to 10, 50 and 100 dimensions. The purpose of this is to see how the EKI algorithm performs when the number of dimensions increases.

### Standard EKI

```{r}
num_dimensions <- c(10, 50, 100)
for (dimension in num_dimensions) {
  num_particles <- 4*dimension
  true_parameters <- list(alpha = 2, sigma = 2, 
                          x = rep(1, dimension), alpha.sd = 5, sigma2.sd = 2)
  eki_result <- eki_normal(num_particles, true_parameters)
  plot_eki_normal_known_var(eki_result, true_parameters)
}
```

### Adaptive Temperature

```{r}
num_dimensions <- c(10, 50, 100)
for (dimension in num_dimensions) {
  num_particles <- 4*dimension
  true_parameters <- list(alpha = 2, sigma = 2, 
                          x = rep(1, dimension), alpha.sd = 5, sigma2.sd = 2)
  eki_result <- eki_normal_adaptive(num_particles, true_parameters)
  plot_eki_normal_known_var(eki_result, true_parameters)
}
```

## Changing $\alpha$

In this experiment, I keep $\sigma^2$ fixed at 4, make $x$ a 50-dimensional vector of 1s and change $\alpha$ from 0 to 10 in increments of 2. The purpose of this is to make sure that the EKI algorithm is accurate across of range of $\alpha$ and if it changes how the noise parameter is estimated.

```{r}
alpha_seq <- seq(0, 10, length.out = 6)
```

### Standard EKI

```{r}
num_dimensions <- 50
num_particles <- 4*num_dimensions
for (alpha in alpha_seq) {

  true_parameters <- list(alpha = alpha, sigma = 2, 
                          x = rep(1, num_dimensions), alpha.sd = 5, 
                          sigma2.sd = 2)
  eki_result <- eki_normal(num_particles, true_parameters)
  plot_eki_normal_known_var(eki_result, true_parameters)
}
```

## Changing $\sigma^2$

```{r}
sigma_seq <- seq(2, 10, length.out = 5)
```

```{r}
num_dimensions <- 50
num_particles <- 4*num_dimensions
for (sigma in sigma_seq) {

  true_parameters <- list(alpha = 2, sigma = sigma, 
                          x = rep(1, num_dimensions), alpha.sd = 5, 
                          sigma2.sd = 2)
  eki_result <- eki_normal(num_particles, true_parameters)
  plot_eki_normal_known_var(eki_result, true_parameters)
}
```

## Changing $x$

In this experiment I keep $\alpha$ and $\sigma^2$ fixed at 2 and 4 respectively and let $x$ be a 50 dimensional random vector.

### Standard EKI

```{r}
num_dimensions <- 50
x <- rnorm(50, mean = 1, sd = 1)
true_parameters <- list(alpha = 2, sigma = 2, 
                        x = x, alpha.sd = 5, 
                        sigma2.sd = 2)
eki_result <- eki_normal(num_particles, true_parameters)
plot_eki_normal_known_var(eki_result, true_parameters)
```
