## Import Functions

### External Packages

```{r}
pacman::p_load(pacman, purrr, mvtnorm, mcmc, MASS)
```

### Algorithms

```{r}
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/eki.R')
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/mcmc_normal.R')
```

### Models

```{r}
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/models/eki_normal.R')
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/models/eki_normal_known_var.R')
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/models/eki_normal_known_mean.R')
```

### Sampling and PDFs

```{r}
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/pdfs/pdfs_normal.R')
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/samples/samples_normal.R')
```

### Utils

```{r}
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/utils/tempering.R')
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/utils/eki_helper.R')
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/results/plots_normal.R')
```

```{r}
generate_results <- function(num_particles, true_parameters) {
  
  true_data <- likelihood_normal(true_parameters)
  
  # Model 1
  eki_result_known_var <- eki_normal_known_var(num_particles, true_data, true_parameters, 
                                               prior_params, adaptive = adaptive)
  plot_eki_normal_known_var(eki_result_known_var, true_parameters, prior_params)
  
  # Model 2
  eki_result_known_mean <- eki_normal_known_mean(num_particles, true_data, true_parameters, 
                                                 prior_params, adaptive = adaptive)
  plot_eki_normal_known_mean(eki_result_known_mean, true_parameters, prior_params)
  
  # Model 3
  eki_result <- eki_normal(num_particles, true_data, true_parameters, 
                           prior_params, adaptive = adaptive)
  plot_eki_normal(eki_result, true_parameters, prior_params)
  
  # MCMC Comparison for Model 3
  run.1 <- normal_mcmc(true_data, true_parameters, prior_params)
  chain.1 <- run.1$batch
  run.2 <- normal_mcmc(true_data, true_parameters, prior_params)
  chain.2 <- run.2$batch
  plot_mcmc_trace_plots(chain.1, chain.2, 500)
  plot_mcmc_histogram(chain.1, chain.2, true_parameters, 500)
  
}
```

# Results For Multivariate Normal Model

## The Model

We assume that we have access to a single observation from a multivariate normal distribution $y \sim N(\alpha x, \sigma^2I)$ where $\alpha$ is a scalar and $x$ is a known vector.

I look at three variations of this model

-   The variance is known, but the mean is not $\theta = \alpha$

-   The mean is known, but the variance is not $\theta = \sigma^2$

-   Both parameters are unknown $\theta = (\alpha, \sigma^2)$.

I look at a range of combinations of $\alpha, x$ and $\sigma^2$, comparing against MCMC sampling. Of particular interest is how well GEKI is able to estimate the noise parameter, since previous EKI algorithms require the noise parameter to be known.

### Priors

In all combinations we draw from the same priors (if that parameter is considered unknown):

-   $\alpha \sim ~ N(0, 5^2)$

-   $\log(\sigma^2) \sim N(2, 1^2)$

```{r}
prior_params <- list(alpha.mean = 0, alpha.sd = 5, 
                     sigma2.mean = 2, sigma2.sd = 1)
```

### Adaptive Tempering

```{r}
adaptive = TRUE
```

### Output

For each variation of the model, I plot the EKI histograms of the particles against their prior. In the known mean, unknown variance case, I also plot the posterior since there is an analytical posterior in this special case. This serves as a check that the algorithm has been implemented correctly.

## Changing Number of Dimensions

In this experiment I keep $\alpha$ and $\sigma^2$ fixed at 2 and 4 respectively and let $x$ be a vector of 1s corresponding to 10, 50 and 100 dimensions. I keep the number of particles fixed at 400. The purpose of this is to see how the EKI algorithm performs when the number of dimensions increases.

```{r}
num_dimensions <- c(10, 50, 100)
num_particles <- 400
for (dimension in num_dimensions) {
  true_parameters <- list(alpha = 2, sigma = 2, x = rep(1, dimension))
  generate_results(num_particles, true_parameters)
}
```

## Changing $\alpha$

In this experiment, I keep $\sigma^2$ fixed at 4, make $x$ a 50-dimensional vector of 1s and change $\alpha$ from 0 to 10 in increments of 2. The purpose of this is to make sure that the EKI algorithm is accurate across of range of $\alpha$ and if it changes how the noise parameter is estimated.

```{r}
alpha_seq <- seq(0, 10, length.out = 6)
num_dimensions <- 50 
num_particles <- 400 
for (alpha in alpha_seq) {    
  true_parameters <- list(alpha = alpha, sigma = 2, x = rep(1, num_dimensions))   
  generate_results(num_particles, true_parameters)
}
```

## Changing $\sigma^2$

In this experiment I change the true value of $\sigma$ from $2$ to $10$ in increments of $2$. The purpose of this is to determine whether the EKI algorithm can detect values of $\sigma$ which are further away from the prior. Additionally, we would expect the $\alpha$ particles to become more diffuse as the true value of $\sigma$ increases.

```{r}
sigma_seq <- seq(2, 10, length.out = 5)
num_dimensions <- 50
num_particles <- 400
for (sigma in sigma_seq) {

  true_parameters <- list(alpha = 2, sigma = sigma, x = rep(1, num_dimensions))
  generate_results(num_particles, true_parameters)
}
```

## Changing Number of Particles

```{r}
num_dimensions <- 50
particle_seq <- c(100, 500, 1000) 
true_parameters <- list(alpha = 2, sigma = 2, x = rep(1, num_dimensions))

for (num_particles in particle_seq) {
  generate_results(num_particles, true_parameters)
}
```
