---
output: 
  github_document
---

```{r}
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/utils/check_covariances.R')
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/models/eki_malaria.R')
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/models/eki_malaria_known_var.R')
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/models/eki_malaria_known_d_in.R')
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/models/eki_malaria_d_in_only.R')
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/samples/samples_malaria.R')
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/utils/eki_helper.R')
```

```{r}
plot_covariance_against_sigma <- function(particle_func, data_func, 
                                          covariance_func, covariance_type, column,
                                          correlation = F) {
  sigma_seq <- seq(-2, 1, length.out = 10)
  res <- c()
  for (sigma in sigma_seq) {
    prior_params <- list(din.sd = 2,
                         phi.mean = 0, phi.sd = 1,
                         eta0.mean = 0, eta0.sd = 1,
                         sigma.mean = sigma, sigma.sd = 1)
    particles <- particle_func(50, prior_params)
    # print(exp(sigma + 0.5)**2)
    # print(mean(particles[, 4]))
    true_params <- list(sigma = 0.5, phi = 0.25, eta0 = 0.11, d_in = 0.5)
    true_unconstrained_params <- unconstrain_malaria_params(true_params)
    likelihood_samples <- data_func(50, particles, true_unconstrained_params)
    # print(mean(likelihood_samples))
    covariances <- calculate_covariances(particles, likelihood_samples, 
                                         correlation = correlation)
    res <- c(res, covariance_func(covariances, column))
  }
  print(res)
  plot(exp(sigma_seq), res, xlab = expression(sigma), ylab = 'Covariance', 
       main = glue('Empirical {covariance_type}'), cex.main = 2, cex.lab = 1.5)
  abline(h = 0, col = 'red')
}

plot_data_diff_against_sigma <- function(true_data, particle_func, 
                                        data_func, column) {
  
  d_y <- length(true_data)
  # I'm replicating this data for the number of particles to make the dimensions     easier to work with
  simulated_data <- matrix(true_data, nrow = 50, ncol = d_y, byrow = T)
  
  sigma_seq <- seq(-2, 1, length.out = 10)
  res <- c()
  
  for (sigma in sigma_seq) {
    prior_params <- list(din.sd = 2,
                         phi.mean = 0, phi.sd = 1,
                         eta0.mean = 0, eta0.sd = 1,
                         sigma.mean = sigma, sigma.sd = 1)
    particles <- particle_func(50, prior_params)
    true_params <- list(sigma = 0.5, phi = 0.25, eta0 = 0.11, d_in = 0.5)
    true_unconstrained_params <- unconstrain_malaria_params(true_params)
    likelihood_samples <- data_func(50, particles, true_unconstrained_params)
    # print(mean(likelihood_samples))
    res <- c(res, mean(simulated_data - likelihood_samples))
    
  }
  print(res)
  plot(exp(sigma_seq + 1/2), res, xlab = expression(sigma), ylab = 'Covariance', 
       main = 'Movement of Particles', cex.main = 2, cex.lab = 1.5)
  abline(h = 0, col = 'red')
    
  
}

plot_movement_against_sigma <- function(true_data, particle_func, 
                                        data_func, column) {
  
  d_y <- length(true_data)
  # I'm replicating this data for the number of particles to make the dimensions     easier to work with
  simulated_data <- matrix(true_data, nrow = 50, ncol = d_y, byrow = T)
  
  sigma_seq <- seq(-2, 1, length.out = 10)
  res <- c()
  
  for (sigma in sigma_seq) {
    prior_params <- list(din.sd = 2,
                         phi.mean = 0, phi.sd = 1,
                         eta0.mean = 0, eta0.sd = 1,
                         sigma.mean = sigma, sigma.sd = 1)
    particles <- particle_func(50, prior_params)
    true_params <- list(sigma = 0.5, phi = 0.25, eta0 = 0.11, d_in = 0.5)
    true_unconstrained_params <- unconstrain_malaria_params(true_params)
    likelihood_samples <- data_func(50, particles, true_unconstrained_params)
    # print(mean(likelihood_samples))
    covariances <- calculate_covariances(particles, likelihood_samples)
    new_particles <- update_particles(1, particles, simulated_data, 
                                  likelihood_samples, covariances, 50)
    update_size <- new_particles - particles
    res <- c(res, mean(update_size[, column]))
    
  }
  print(res)
  plot(exp(sigma_seq), res, xlab = expression(sigma), ylab = 'Covariance', 
       main = 'Movement of Particles', cex.main = 2, cex.lab = 1.5)
  abline(h = 0, col = 'red')
    
  
}
```

# Checking Covariances

When testing the EKI algorithm on a multivariate normal model with unknown mean and noise, it was found that the EKI algorithm was not able to identify the true noise parameter, and was anchored to the prior distribution. The purpose of this markdown document is to explain this phenomenon.

The main functions for this document can be found in `src/utils/check_covariances.R` I am interested in looking at three variables:

-   How a covariance matrix changes with the prior mean of $\sigma$: `plot_covariance_against_sigma`

-   How the difference between the true data and the samples changes with the prior mean of $\sigma$

-   How the final update size changes with the prior mean of $\sigma$

```{r}
# Can change these functions to fix one of the parameters
initialise_malaria_func <- initialise_malaria_particles
synthetic_malaria_func <- synthetic_malaria
```

# Findings

## Checking $C^{yy}$

```{r}
plot_covariance_against_sigma(initialise_malaria_func, synthetic_malaria_func, 
                              check_c_yy, "Data Variance")
```

As one would expect, when the prior mean of $\sigma$ increases, so does the variance of the data. All else equal, this will have the effect of reducing the step-size.

## Checking $C^{yx}$

```{r}
plot_covariance_against_sigma(initialise_malaria_func, synthetic_malaria_func, 
                              check_c_yx, "cross covariance", 1)
plot_covariance_against_sigma(initialise_malaria_func, synthetic_malaria_func, 
                              check_c_yx, "cross covariance", 2)
plot_covariance_against_sigma(initialise_malaria_func, synthetic_malaria_func, 
                              check_c_yx, "cross covariance", 3)
plot_covariance_against_sigma(initialise_malaria_func, synthetic_malaria_func,
                              check_c_yx, "cross covariance", 4)
```

For the phase shift parameter $\phi$ and $\sigma^2$ we see that the mean of the cross covariances fluctuate around 0. In the case of $\phi$ this makes sense considering $\phi$ will be positively correlated with some parts of the data and negatively correlated with others. In the case of $\sigma^2$ we expect no correlation with the data as in the MVN model.

For $d_{in}$ the covariance actually decreases in magnitude as $\sigma^2$ increases. This is somewhat counter-intuitive, but most likely because $d_{in}$ does not change the mean as significantly as $\eta_0$. Thus, when the variance is higher, any value of $d_{in}$ may provide a reasonable fit to the data.

For $\eta_0$ the covariance fluctuates around a constant value.

```{r}
plot_covariance_against_sigma(initialise_malaria_func, synthetic_malaria_func, 
                              check_c_yx, "cross correlation", 1, 
                              correlation = T)
plot_covariance_against_sigma(initialise_malaria_func, synthetic_malaria_func, 
                              check_c_yx, "cross correlation", 2, 
                              correlation = T)
plot_covariance_against_sigma(initialise_malaria_func, synthetic_malaria_func, 
                              check_c_yx, "cross correlation", 3,
                              correlation = T)
plot_covariance_against_sigma(initialise_malaria_func, synthetic_malaria_func,
                              check_c_yx, "cross correlation", 4,
                              correlation = T)
```

In terms of correlation, all of the correlations go to zero as $\sigma^2$ increases, suggesting the signal between the parameters and the data is getting weaker.

## Checking Stepsize

I am interested in looking at the columns of the matrix $C^{xy}(C^{yy})^{-1}$ which is equivalent to $C^{xy}(C^{yy} +(h_l^{-1}-1)C^{y|x})^{-1}$ when $h_l = 1$ (i.e. no tempering)

```{r}
plot_covariance_against_sigma(initialise_malaria_func, synthetic_malaria_func, 
                              check_stepsize, "Stepsize", 1)
plot_covariance_against_sigma(initialise_malaria_func, synthetic_malaria_func,
                              check_stepsize, "Stepsize", 2)
plot_covariance_against_sigma(initialise_malaria_func, synthetic_malaria_func,
                              check_stepsize, "Stepsize", 3)
plot_covariance_against_sigma(initialise_malaria_func, synthetic_malaria_func,
                              check_stepsize, "Stepsize", 4)
```

For all the particles, the stepsize is quite small, although there is not a clear pattern as $\sigma^2$ increases.

## Checking Data Difference

```{r}
true_params <- list(sigma = 0.5, phi = 0.25, eta0 = 0.11, d_in = 0.5)
true_unconstrained_params = unconstrain_malaria_params(true_params)
true_data <- likelihood_malaria(true_unconstrained_params)
plot_data_diff_against_sigma(true_data, initialise_malaria_func, 
                            synthetic_malaria_func, 1)
plot_data_diff_against_sigma(true_data, initialise_malaria_func, 
                            synthetic_malaria_func, 2)
plot_data_diff_against_sigma(true_data, initialise_malaria_func, 
                            synthetic_malaria_func, 3)
plot_data_diff_against_sigma(true_data, initialise_malaria_func, 
                            synthetic_malaria_func, 4)
```

For all particles, the mean difference between the true data and likelihood samples is roughly constant as $\sigma^2$ increases, which makes logical sense.

## Checking Update Size

```{r}
true_params <- list(sigma = 0.5, phi = 0.25, eta0 = 0.11, d_in = 0.5)
true_unconstrained_params = unconstrain_malaria_params(true_params)
true_data <- likelihood_malaria(true_unconstrained_params)
plot_movement_against_sigma(true_data, initialise_malaria_func, 
                            synthetic_malaria_func, 1)
plot_movement_against_sigma(true_data, initialise_malaria_func, 
                            synthetic_malaria_func, 2)
# plot_movement_against_sigma(true_data, initialise_malaria_func, 
#                             synthetic_malaria_func, 3)
# plot_movement_against_sigma(true_data, initialise_malaria_func, 
#                             synthetic_malaria_func, 4)
```

# Implications
