---
title: "Generalised Ensemble Kalman Inversion - Misspecified Noise"
author: "Owen Jackson"
date: "2025-01-15"
output: github_document
---

# Results for Misspecified Noise

## Imports

### External Functions

```{r}
pacman::p_load(pacman, testthat, deSolve, rootSolve, MASS, mvtnorm, extraDistr, 
               purrr, glue, logitnorm)
```

### Algorithms

```{r}
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/eki.R')
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/eki_known_noise.R')
```

### Models

```{r}
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/models/eki_malaria.R')
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/models/eki_malaria_known_var.R')
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/models/eki_malaria_known_d_in.R')
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/models/eki_malaria_d_in_only.R')
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/models/eki_normal_known_var.R')
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/models/eki_malaria_var_only.R')
```

### **Sampling and PDFS**

```{r}
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/pdfs/pdfs_normal.R')
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/samples/samples_normal.R')
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/pdfs/pdfs_malaria.R')
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/samples/samples_malaria.R')
```

### Utils

```{r}
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/utils/eki_helper.R')
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/src/utils/tempering.R')
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/results/plots_normal.R')
source('C:/Users/owenj/OneDrive/Uni/Vacation Scholarship/GEKI_Vacation_Scholarship/results/plots_malaria.R')
```

```{r}
adaptive = TRUE
```

## Introduction

When looking at the multivariate normal model, it was found that GEKI does not do a good job at estimating the noise parameter. Following on from those results, the next step of the project was to investigate whether GEKI outperforms EKI in the cases where the noise parameter is known, but misspecified. Specifically, it is chosen to be too small. In the context of EKI we would expect the concentration of the particles to be too narrow and potentially away from the true parameter when we diverge from the linear Gaussian setting.

## Malaria Transmission Model

The observational model is $y(t) \sim LogNormal(log(W(t)-W(t-1)), \sigma^2)$ where $W(t)$ is the cumulative number of malaria cases by time $t$.

Unlike when we work in log space, $\sigma$ will effect the mean of the observations, and thus we anticipate that EKI will be able to identify $\sigma$.

```{r}
num_particles <- 400

prior_params <- list(din.sd = 2,
                     phi.mean = 0, phi.sd = 1,
                     eta0.mean = 0, eta0.sd = 1)

true_params <- list(sigma = 0.5, phi = 0.25, eta0 = 0.11, d_in = 0.5)
true_unconstrained_params = unconstrain_malaria_params(true_params)
true_data <- likelihood_malaria(true_unconstrained_params, log_obs = F)
```

### Four Unknown Parameters

Initially, we tried looking at $\theta = (d_{in}, \phi, \eta_0, \sigma)$

```{r}
eki_result <- eki_malaria(num_particles, true_data, true_unconstrained_params, 
                          prior_params, adaptive = adaptive)
plot_eki_malaria(eki_result, true_params, prior_params)
```

However, when the prior mean for $\sigma$ was too large, the algorithm was unable to identify any of the parameters. Essentially, the $\sigma$ parameter becomes the dominant parameter for the mean, and the EKI algorithm is not able to respond sufficiently to reduce the $\sigma$ parameter enough. **Further investigation is needed to explain this phenomenon.**

When we choose a prior mean for $\sigma$ that was small, we found similar results to the three unknown parameter case (see below). That is, there was still an issue identifying $d_{in}$ and $\eta_0$. We also saw this when looking at log space.

### Three Unknown Parameters

Next, we tried looking at $\theta = (d_{in}, \phi, \eta_0)$ as our unknown parameters.

```{r}
# EKI
eki_result <- eki_malaria_known_var(num_particles, true_data, 
                                    true_unconstrained_params, prior_params, 
                                    general = F, adaptive = adaptive)
plot_eki_malaria_known_var(eki_result, true_params)
plot_eki_posterior_predictive(eki_result, true_data, true_unconstrained_params)
```

```{r}
# GEKI
eki_result <- eki_malaria_known_var(num_particles, true_data, 
                                    true_unconstrained_params, 
                                    prior_params, adaptive = adaptive)
plot_eki_malaria_known_var(eki_result)
plot_eki_posterior_predictive(eki_result, true_data, true_unconstrained_params)
```

While the posterior predictive fits the true data well, the EKI and GEKI algorithms have a difficult time identifying the true $d_{in}$ and $\eta_0$. In that sense, we get values for $d_{in}$ which don't make sense in the context of the model.

```{r}
d_in_particles <- exp(eki_result$particles[, 1]) + 0.16
eta0_particles <- plogis(eki_result$particles[, 3])
plot(eta0_particles, d_in_particles)
model <- lm(d_in_particles ~ eta0_particles)
abline(model, col = 'red')
cor(d_in_particles, eta0_particles)
```

Upon closer inspection we see a strong positive correlation between the $d_{in}$ and $\eta_0$ particles. Essentially the higher values offset each other, producing predictions which fit the true data. The reason this has more of an effect on $d_{in}$ is most likely a combination of the prior distributions and because $d_{in}$ is more sensitive to changes in $\eta_0$.

It is worth noting that this co-identifiability issue is not unique to EKI; similar results were found when running SMC on the same model.

### Two Unknown Parameters

Based on the above findings, it was decided to instead look at $\theta = (\eta_0, \phi)$.

```{r}
# EKI
eki_result <- eki_malaria_known_d_in(num_particles, true_data, 
                                    true_unconstrained_params, prior_params, 
                                    general = F, adaptive = F)
plot_eki_malaria_known_d_in(eki_result, true_params, prior_params)
plot_eki_posterior_predictive_known_d_in(eki_result, true_data, true_unconstrained_params)
```

```{r}
# GEKI
eki_result <- eki_malaria_known_d_in(num_particles, true_data, 
                                    true_unconstrained_params, prior_params, 
                                    adaptive = adaptive)
plot_eki_malaria_known_d_in(eki_result, true_params, prior_params)
plot_eki_posterior_predictive_known_d_in(eki_result, true_data, true_unconstrained_params)
```

### Misspecified Noise

We again consider $\theta = (\eta_0, \phi)$, but under-specify the noise, relative to the true data. Since the noise of the true data is $\sigma = 0.5$ we choose instead $\sigma = 0.33$.

```{r}
true_params <- list(sigma = 0.33, phi = 0.25, eta0 = 0.11, d_in = 0.5)
true_unconstrained_params = unconstrain_malaria_params(true_params)
```

```{r}
# GEKI
eki_result <- eki_malaria_known_d_in(num_particles, true_data, 
                                    true_unconstrained_params, prior_params, 
                                    adaptive = adaptive)
plot_eki_malaria_known_d_in(eki_result, true_params, prior_params)
plot_eki_posterior_predictive_known_d_in(eki_result, true_data, true_unconstrained_params)
```

### $d_{in}$ Only

```{r}
eki_result <- eki_malaria_d_in_only(num_particles, true_data, true_unconstrained_params, general = F, 
                                    prior_params, adaptive = adaptive)
plot_eki_malaria_d_in_only(eki_result, true_params)
```

### $\sigma$ Only

As we noted above, the algorithm has a difficult time identifying four parameters. We now try $\theta = \sigma$.

```{r}
eki_result <- eki_malaria_var_only(num_particles, true_data, 
                                    true_unconstrained_params, prior_params, 
                                    adaptive = F)
plot_eki_malaria_var_only(eki_result, true_params, prior_params)
```

In this instance, we find that if the prior mean for $\sigma$ is set low enough that the algorithm is able to concentrate around the true value. However, when we set the prior mean too high, the variance blows up and we get back the prior distribution. In practice, this is concerning for the usefulness of the algorithm, because we shouldn't be required to massage the priors to get good results.
